---
title: "STA207 Project2 Code"
author: "Yahui Li"
date: "2020/1/28"
output: 
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r,message=FALSE}
library(tidyverse)
library(stats)
library(car)
library(ggplot2)
library(kableExtra)
library(gridExtra)
library(ggpubr)
```

1.  Explore math scaled scores in the 1st with teachers as the unit. Generate summary statistics (in forms of tables or plots) that you find informative, and explain them. 

```{r}
# https://dataverse.harvard.edu/file.xhtml?fileId=666716&version=1.0
load("./STAR_Students.RData")
data <- data.frame(g1classtype = x$g1classtype, g1schid = x$g1schid, 
                   g1tchid = x$g1tchid, g1tmathss = x$g1tmathss)
```

```{r}
# Table code
varsum <- data.frame(Name= c("g1classtype", "g1schid", "g1tchid", "g1tmathss"),
                     Type= c("qualitative variable", "qualitative variable", "qualitative variable", "quantitive variable"), 
                     "The number of levels"= c(3,76,339,NA))
kable(varsum,caption = "Table 1: basic information of variables") %>%
  kable_styling(bootstrap_options = "striped", full_width = F )
```

In Figure 1, the above plot shows the percentage of missing value of four variables are nearly same, and the plot below shows that the location of missing values for three qualitative variables are consistent. It can be explained that because of low grades in kindergarten in STAR project, students quit STAR project and go to another school, or they join STAR project after 1st grade. Because we don't know how to tackle missing values, considering that we have enough data and believe that it won't have the significant impact on the results, we drop them directly.

```{r}
#Code for the first plot
missing.values <- data %>%
  gather(key = "key", value = "val") %>%
  mutate(isna = is.na(val)) %>%
  group_by(key) %>%
  mutate(total = n()) %>%
  group_by(key, total, isna) %>%
  summarise(num.isna = n()) %>%
  mutate(pct = num.isna / total * 100)

levels <-
    (missing.values  %>% filter(isna == T) %>% arrange(desc(pct)))$key

percentage.plot <- missing.values %>%
      ggplot() +
        geom_bar(aes(x = reorder(key, desc(pct)), 
                     y = pct, fill=isna), 
                 stat = 'identity', alpha=0.8) +
      scale_x_discrete(limits = levels) +
      scale_fill_manual(name = "", 
                        values = c('steelblue', 'tomato3'), labels = c("Present", "Missing")) +
      coord_flip() +
      labs(title = "Figure 1: Percentage of missing values", x =
             'Variable', y = "% of missing values")

#Code for the second plot
row.plot <- data %>%
  mutate(id = row_number()) %>%
  gather(-id, key = "key", value = "val") %>%
  mutate(isna = is.na(val)) %>%
  ggplot(aes(key, id, fill = isna)) +
    geom_raster(alpha=0.8) +
    scale_fill_manual(name = "",
        values = c('steelblue', 'tomato3'),
        labels = c("Present", "Missing")) +
    scale_x_discrete(limits = levels) +
    labs(x = "Variable",
           y = "Row Number", title = "Missing values in rows") +
    coord_flip()


#arrange them
grid.arrange(percentage.plot, row.plot, nrow = 2)
```

```{r}
#remove NA
data <- na.omit(data)
data$g1schid <- factor(data$g1schid)
data$g1tchid <- factor(data$g1tchid)
```


In this project, we are interested in math scaled scores in the 1st grade with teachers as the unit. After verifying each teacher are only employed in specific school and teach one specific class type, we select the mean of math scaled scores of one particular class to judge teachers. Therefore, we have a new dataset which has 339 rows and four columns including g1classtype, g1schid, g1tchid and mean. Figure 2 shows the distribution of the mean of math score by different teacher. It is bell-shaped and roughly closed to normal distribution.

```{r}
data_by_teacher <- data %>%
  group_by(g1classtype,g1schid,g1tchid) %>%
  summarize(mean = mean(g1tmathss))

ggplot(data_by_teacher, aes(x=mean)) + 
  geom_histogram() +
  ggtitle("Figure 2: the mean of math score by different teacher")

```

Figure 3 shows the frequency number of class type and school id. Due to too many levels in g1schid, x axis labels in the plot of g1schid is omitted for readability. We can find that  g1classtype data is roughly balanced, and there is one specific school whose frequency number which is much higher than rest of schools. Figure 4 shows that the frequency number of class of different types in each school. Excepts a few extremely cases, data is more balanced than data group by class type. 

```{r}
# Figure 3
classbar <- ggplot(data_by_teacher, aes(g1classtype)) +
  geom_bar(fill = "#0073C2FF") +
  theme_pubclean() +
  ggtitle("Figure 3: The frequency bar plots of class type and school id")


schoolbar <- ggplot(data_by_teacher, aes(x=reorder(g1schid, -table(g1schid)[g1schid]))) +
  geom_bar(fill = "#0073C2FF") +
  theme_pubclean() +
  theme(
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(x = "g1schid")


grid.arrange(classbar, schoolbar, nrow = 2) 
```

```{r}
# Figure 4 Bar plots of contingency tables
ggplot(data_by_teacher %>%
  group_by(g1classtype, g1schid) %>%
  count(), aes(x = g1schid, y = n))+
  geom_bar(
    aes(fill = g1classtype), stat = "identity", color = "white",
    position = position_dodge(0.9)) +
  theme(
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(x = "g1schid") +
  ggtitle("Figure 4: The frequency bar plots of grouped contingency tables")

```

In Figure 5, it can be seen that different class types have effects on mean scores of math. Therein, teachers in small class have the higher mean scores of math and teachers in regular class have the lower mean scores of math in general. Also, different schools have impacts on mean scores of math. Aming to investigate the class types effect on mean scores of math, it is a good strategy that using schools to creat blocks for eliminating effects between different schools. 

```{r}

classbox <- ggplot(data_by_teacher, aes(x=g1classtype, y=mean)) + 
    geom_boxplot(fill="slateblue", alpha=0.2) +
    labs(title = "Figure 5: The distribution of mean scores of math by class type and school id")


schoolbox <- ggplot(data_by_teacher, aes(x=g1schid, y=mean)) + 
    geom_boxplot(fill="slateblue", alpha=0.2) +
    theme(
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(x = "g1schid") 

grid.arrange(classbox, schoolbox, ncol = 2) 
```


We can check each teacher ID was in only one certain type of class and in only one school ID, so we can summarize the sore by different teacher ID.

```{r}
(data_by_teacher <- data %>%
  group_by(g1classtype,g1schid,g1tchid) %>%
  summarize(mean = mean(g1tmathss), sd = sd(g1tmathss), median = median(g1tmathss),
            min = min(g1tmathss),max = max(g1tmathss)))
par(mfrow=c(2,3))
hist(data_by_teacher$mean, xlab = c("mean of math score"), 
     main =title("Histogram of mean"))
hist(data_by_teacher$sd, xlab = c("sd of math score"), 
     main =title("Histogram of standard deviation"))
hist(data_by_teacher$median, xlab = c("median of math score"), 
     main =title("Histogram of median"))
hist(data_by_teacher$min, xlab = c("min of math score"), 
     main =title("Histogram of min"))
hist(data_by_teacher$max, xlab = c("max of math score"), 
     main =title("Histogram of max"))
hist(data_by_teacher$max-data_by_teacher$min, xlab = c("range of math score"), 
     main =title("Histogram of range"))
par(mfrow=c(1,1))
```

From the histogram on the mean of math score by different teacher is normal-like. However, the histogram on standard deviation of math score by different teacher is thiner than normal distribution.



3.  Explain why your model is appropriate for this task on this data set. You may want to include statistics and plots in your explanation. 

```{r}
pairs(data_by_teacher[c(4,1,2)])
```

From the pairwise plot there is no non-linearity pattern. 

```{r,warning= FALSE}
interaction.plot(data_by_teacher$g1classtype, data_by_teacher$g1schid, 
                 data_by_teacher$mean,legend = FALSE)
```

4.  Fit the model you choose in Task 2 and show your fits in the report. 


```{r}
full_model=lm(mean~g1classtype+g1schid+g1classtype*g1schid,data=data_by_teacher);

reduced_model=lm(mean~g1classtype+g1schid,data=data_by_teacher);

anova(reduced_model,full_model)
```

The test result show that interaction effects are very likely to be absent from this data set. This means that we need to treat each combination as a unit, whereas we can compare each type of main effects separately.


5.  Conduct model diagnostic and/or sensitivity analysis. 

a. independence, need descriptive 

In this experiment, researchers assigned students to small classes, regular classed and regular classes with aide randomly.

From the Residuals vs Fitted Values scatterpoint, we can see that the residuals are around zero. It shows that the zero mean assumption.

Besides, these points are uniformly distributed on both sides of x-axis, which means that our model does not violate the equal variance assumption.

```{r,warning=FALSE}
anova.fit<-aov(mean~g1classtype+g1schid,data=data_by_teacher)
#check the mean zero and equal variance
plot(anova.fit, which = 1)
plot(anova.fit, which = 2)
summary(anova.fit)
```

The Normal Q-Q plot illustrates that the residuals are a little bit heavy tailed compared with the normal distribution. In order to check the normality assumption more precisely, we conduct the Shapiro-Wilk Normality Test. The result shows that $W=0.9802$ and the $p-value$ is 0.0001, which is much smaller than the significant level 0.01. Therefore, we are 99% confident that the residuals are normally distributed and our model follows the normality assumption.

```{r}
#normality assumption
plot(anova.fit, which = 2)

shapiro.test(anova.fit$residuals)
```

```{r}
# Krusal-Wallis test:
kruskal.test(mean ~ interaction(g1classtype, g1schid) , data = data_by_teacher)
```

```{r}
# Krusal-Wallis test:

kruskal.test(mean ~ interaction(g1classtype, g1schid) , data = data_by_teacher)
```


6.  Test whether there is a difference in math scaled score in 1st grade across teachers in different class types. Justify your choice of test. 

We can use the Tukey-Kramer method for this task.

```{r}
alpha=0.05;
T.ci=TukeyHSD(anova.fit,conf.level = 1-alpha)
plot(T.ci, las=1 , col="brown")
T.ci$g1classtype
```
