---
title: "STA207 Project2 Code"
author: "Yahui Li"
date: "2020/1/28"
output: 
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r,message=FALSE}
library(tidyverse)
```

1.  Explore math scaled scores in the 1st with teachers as the unit. Generate summary statistics (in forms of tables or plots) that you find informative, and explain them. 

```{r}
# https://dataverse.harvard.edu/file.xhtml?fileId=666716&version=1.0
load("~/myrepo/STA207Project2/STAR_Students.RData")
data <- data.frame(g1classtype = x$g1classtype, g1schid = x$g1schid, 
                   g1tchid = x$g1tchid, g1tmathss = x$g1tmathss)
```

Use no.omit() to remove all NA in dataset

```{r}
data <- na.omit(data)
```

Test the type of each varaible.

```{r}
sapply(data, class)
#We should change ID variables as factor.
data$g1schid <- factor(data$g1schid)
data$g1tchid <- factor(data$g1tchid)
nlevels(data$g1schid) #there are 76 school IDs
nlevels(data$g1tchid) #there are 339 teacher IDs
```

We can check each teacher ID was in only one certain type of class and in only one school ID, so we can summarize the sore by different teacher ID.

```{r}
(data_by_teacher <- data %>%
  group_by(g1classtype,g1schid,g1tchid) %>%
  summarize(mean = mean(g1tmathss), sd = sd(g1tmathss), median = median(g1tmathss),
            min = min(g1tmathss),max = max(g1tmathss)))
par(mfrow=c(2,3))
hist(data_by_teacher$mean, xlab = c("mean of math score"), 
     main =title("Histogram of mean math score"))
hist(data_by_teacher$sd, xlab = c("sd of math score"), 
     main =title("Histogram of standard deviation on math score"))
hist(data_by_teacher$median, xlab = c("median of math score"), 
     main =title("Histogram of median on math score"))
hist(data_by_teacher$min, xlab = c("min of math score"), 
     main =title("Histogram of min on math score"))
hist(data_by_teacher$max, xlab = c("max of math score"), 
     main =title("Histogram of max on math score"))
hist(data_by_teacher$max-data_by_teacher$min, xlab = c("range of math score"), 
     main =title("Histogram of range on math score"))
par(mfrow=c(1,1))
```

From the histogram on the mean of math score by different teacher is normal-like. However, the histogram on standard deviation of math score by different teacher is thiner than normal distribution.



2.  Write down a two-way ANOVA model to study the effects of class types on math scaled scores in 1st grade, with the school indicator as the other factor. Explain your notation.

$$Y_{ijk} = \mu_{ij} + \epsilon_{ijk} \text{ with } \epsilon_{ijk} \overset{\mathrm{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2),$$

where 
$\mu_{ij} = \mu + \alpha_i + \beta_j + \gamma_{ij}$.

By notation, 

$i$ represents the treatment of class type, and  $i = 1, 2, 3$ means the type of class is `small`, `regular` and `regular with aide`.

$j$ represents the treatment of school ID. $j = 1, \cdots, 76$.

$k$ represents summary information about math score by different cases. $k = 1, \cdots, 339$.

$Y_{ijk}$ refers to  

where $\mu$ is the grand mean, $\alpha_i$ is the additive main effect of level $i $ from the first factor (''i''-th row in the contigency table), $\beta_j$ is the additive main effect of level $j$ from the second factor (''j''-th column in the contingency table) and $\gamma_{ij}$ is the non-additive interaction effect of treatment $(i,j)$ from both factors (cell at row ''i'' and column ''j'' in the contingency table).

Another equivalent way of describing the two-way ANOVA is by mentioning that, besides the variation explained by the factors, there remains some [[statistical noise]]. This amount of unexplained variation is handled via the introduction of one random variable per data point, $\epsilon_{ijk}$, called [[Errors and residuals in statistics|error]]. These $n$ random variables are seen as deviations from the means, and are assumed to be independent and normally distributed:




